---
title: "Biking in the Rain"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
###################################################################################################
## Installs any missing packages -- barring system constraints ( will not fix binary dependencies )
###################################################################################################
installMissingPackages <- function(package_name) {
  if (! (package_name %in% rownames(installed.packages())) ) {
    install.packages(package_name, repos="http://cran.rstudio.com/")
  }
}

#################################################
## Loads all required packages in a package list
################################################
loadRequiredPackages <- function(list) {
  for (package in list) {
    installMissingPackages(package)
    library(package, character.only=TRUE)
  }
}

packages <- c('jsonlite', 'sqldf','dplyr', 'glue', 'bigrquery', 'DBI', 'lubridate', 'ggmap')

loadRequiredPackages(packages)
```

## Summer Bicycle Commuting -- Weather Dependent? 

As I enjoyed my cool, rainy bike ride to Statistics class I wondered to myself if other people feel the same way. Is your average commuter cyclist a fair-weather rider, or do they throw on a rain coat and go ride through the puddles? I realized I may be able to answer my question when I passed the Capitol City Eco-Totem, which counts bike traffic in both directions on the path.  

### Data Qualifications

If I am to answer whether or not bicycle commuters were resilient to the precipitation, I wanted to subset for a few environmental variables.  

- Subset for months when school wasn't in session (July, August).
- Subset for months that had reasonably good weather throughout (July, August).
- Subset for days where folks typically work (M-F).

This led me to selecting the months of July and August when the majority of college students had left town, and the weather was typically good. 

### Data Sources
Dane County has 2 bike path counters which are operated by an organization called eco-public. From scraping the JSON that feeds their website, I was able to get two dataframes which include a date, and count for total bikes that crossed the counter. I combined these two dataframes into one. 

I sourced my weather data from the NOAA station located at the Dane County Regional Airport. I selected years 2015 - 2018.
 
### Bike Path Counters and Location
http://www.eco-public.com/public2/?id=100020865  
http://www.eco-public.com/public2/?id=100016754

```{r locate_bike_paths, warning=FALSE, echo=FALSE}
eco_totems <- data.frame(lat=c(43.067913,43.06741), lon=c(-89.41238,-89.38508))
map <- get_googlemap("madison wisconsin", zoom = 14, maptype='terrain')
p <- ggmap(map)
p + geom_point(data=eco_totems, aes(x=lon, y=lat),size=5, col=I("blueviolet"))
```

```{r get_data, warning=FALSE, echo=FALSE}
# Set query date as yesterday
# query_date <- gsub("-", "", (Sys.Date()-2))

query_date <- "20180806"

# Gather Bike Path Data
# The Capitol City Path didn't begin counting until June 2015. We want to start our analysis is May. So we'll do May 2015 onwards until today
southwest_path <- fromJSON(glue::glue("http://www.eco-public.com/api/cw6Xk4jW4X4R/data/periode/100016754?begin=20150630&end={query_date}&step=4"))
capitol_city <- fromJSON(glue::glue("http://www.eco-public.com/api/cw6Xk4jW4X4R/data/periode/100020865?begin=20150630&end={query_date}&step=4"))

# Only use May thru October
southwest_path <- southwest_path %>% 
                    select(date,comptage) %>% 
                    filter(month(date) %in% c(07,08))
# race condition where we receive a duplicate on 2016-10-30 that is null
southwest_path <- na.omit(southwest_path)

capitol_city <- capitol_city %>% 
                    select(date,comptage) %>% 
                    filter(month(date) %in% c(07,08))
# race condition where we receive a duplicate on 2016-10-30 that is null
capitol_city <- na.omit(capitol_city)

bike_path_counts <- sqldf("select southwest_path.date, SUM(southwest_path.comptage + capitol_city.comptage) AS count from southwest_path join capitol_city ON southwest_path.date = capitol_city.date GROUP BY southwest_path.date")

# race condition where we receive a duplicate on 2016-10-30 that is null

# BigQuery connection
# billing <- 'passive-capture'
# 
# con <- dbConnect(
#   bigrquery::bigquery(),
#   project = "bigquery-public-data",
#   dataset = "noaa_gsod",
#   billing = billing
# )
# con

## BigQuery SQLdb
# SELECT 
#   PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da)) AS date,
#   temp,
#   max AS temp_max,
#   min AS temp_min,
#   prcp
# FROM 
# `bigquery-public-data.noaa_gsod.gsod201*` gsod
# JOIN `bigquery-public-data.noaa_gsod.stations` stations
# ON gsod.stn=stations.usaf
# AND gsod.wban=stations.wban
# WHERE state = 'WI'
# AND EXTRACT(MONTH FROM PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da))) > 6
# AND EXTRACT(MONTH FROM PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da))) < 9
# AND stations.name = 'DANE CO REGIONAL-TRUAX FIELD'
# AND _TABLE_SUFFIX BETWEEN '5' and '8'
# ORDER BY date;

# sql_query <- glue::glue("SELECT PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da)) AS date, temp, max AS temp_max, min AS temp_min, prcp FROM `bigquery-public-data.noaa_gsod.gsod201*` gsod JOIN `bigquery-public-data.noaa_gsod.stations` stations ON gsod.stn=stations.usaf AND gsod.wban=stations.wban WHERE state = 'WI' AND EXTRACT(MONTH FROM PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da))) > 6 AND EXTRACT(MONTH FROM PARSE_TIMESTAMP('%Y-%m-%d',CONCAT(gsod.year, '-', gsod.mo, '-', gsod.da))) < 9 AND stations.name = 'DANE CO REGIONAL-TRUAX FIELD' AND _TABLE_SUFFIX BETWEEN '5' and '8' ORDER BY date;")
# 
# weather_data <- dbGetQuery(con, sql_query)

weather_data <- read.csv('weather_data.csv')

bike_path_counts$date <- as.Date(bike_path_counts$date)
weather_data$date <- as.Date(weather_data$date)


df <- merge(bike_path_counts,weather_data)

df <- df %>%
        select(date,count,temp,temp_min,temp_max,prcp) %>% 
        filter(!wday(date) %in% c(1, 7))
```
<br/>
**Data Used for Analysis:**  
```{r head_datum, echo=FALSE}
print(head(df))
```  
### Question  
Is there a statistically significant relationship between the number of Madison commuters counted on our bike paths relative to the amount of precipitation on a given weekday in July & August?  

**Hø:** There is no relationship between precipitation and commuter counts  

**Hₐ:** There is a relationship between precipitation and commuter counts   
### Assumptions
First step is to visualize our populations and determine whether we have a large enough sample, and whether our sample population is normally distributed.  
```{r normal_plots, echo=FALSE}
par(mfrow=c(2,2))
qqnorm(southwest_path$comptage, main='Q-Q Plot for SW Path Counts')
qqnorm(capitol_city$comptage, main='Q-Q Plot for CC Path Counts')
qqnorm(df$count, main='Q-Q Plot for Total Bike Counts')
qqnorm(df$prcp, main='Q-Q Plot for Precipitation')
```

In light of the Q-Q plots that were generated from our data, we can assume normal distributions for all of the bike count data -- however it looks like our precipitation sample is not normal as it is left skewed.  

During my exploration, it was also apparent that there is a very high variance in daily commuter numbers for our 158 observations.   

**s = 987.17**  
```{r variance_in_commuters, echo=FALSE}
plot(df$count,col = 1 + (df$prcp > 0), ylim=c(0,7000), main='High Variance in Commuters', ylab='Commuter Count', xlab='By Day')
abline(h=mean(df$count))
text(10, 200+mean(df$count), "x̅", col = "black")
text(25,100, glue::glue("s = {sd(df$count)}"))
```

This high variance is also represented in each individual bike path counter, so I will assume this variance is not due to a faulty operation of one of the path counters. Or perhaps long term construction that had one of the counters offline.      
<br/>
**s(capitol_city) = 665.70**  
**s(southwest_path) = 441.26**

```{r similar_variance_in_individual_locations, echo=FALSE}
par(mfrow=c(1,2))
plot(capitol_city$comptage,col = 1 + (df$prcp > 0), ylim=c(0,5000), main='Capitol City Commuters', ylab='Commuter Count', xlab='By Day')
abline(h=mean(capitol_city$comptage))
text(15, 200+mean(capitol_city$comptage), "x̅", col = "black")

plot(southwest_path$comptage,col = 1 + (df$prcp > 0), ylim=c(0,5000), main='Southwest Path Commuters', ylab='Commuter Count', xlab='By Day')
abline(h=mean(southwest_path$comptage))
text(15, 200+mean(southwest_path$comptage), "x̅", col = "black")
```

### Regression

With the awareness that my precipitation distribution is not normal on account of the majority of data points being 0, I'm going to do a quick regression anyways -- mostly for fun. 

```{r linear_regression_on_bp_v_prcp}
summary(lm(df$count~df$prcp))
```

```{r show_linear_model, echo=FALSE}
plot(x=df$prcp, y=df$count, ylim=c(0,7000), main='Regression Line', ylab='Biker Count', xlab='Precipitation')
abline(lm(df$count~df$prcp), col='blueviolet')
```

### Results
In light of our regression analysis, it appears there is a statistically significant relationship between commuter counts and precipitation on any given day in July or August. 
<br/>
**p = .00125**
<br/>
Since our p-value is less than .05, we could cautiously consider rejecting **Hø**, however since our precipitation data was not from a normal population I can't fully support this significance. It is interesting nonetheless, and gret exploration.  

### T-Test
Since our precipiation sample is not distributed normally and our ridership counts are, this is the perfect example to subset the commuter counts data into two sets (precipitation/non-precipitation) and perform a Welch's T-Test. This should enable us to determine a relationship between the mean commuter counts on rainy days versus non rainy days. I have subsetted the data into two groups, with one being 49 counts on days of precipitation, and the other being 49 counts on days of no precipitation.  
```{r reformat_data, echo=FALSE}
rain_counts <- df %>%
                 select(count, prcp) %>%
                 filter(prcp > 0)
non_rain <- counts <- df %>%
                     select(count, prcp) %>%
                     filter(prcp == 0) %>%
                     slice(1:49)
```
### Question  
Is there a statistically significant relationship between the number of Madison commuters counted on our bike paths relative to the amount of precipitation on a given weekday in July & August?  

**Hø:** μ(commuter_counts_with_precip) - μ(commuter_counts_no_precip) = 0  

**Hₐ:** μ(commuter_counts_with_precip) - μ(commuter_counts_no_precip) ≠ 0   
```{r visualize_t}
boxplot(non_rain$count, rain_counts$count, ylim=c(0,7000),names=c("No Precip","Precip"), col=c('blanchedalmond','aquamarine3'), ylab='Commuter Count')
```
<br/>
### Assumptions
Let's examine the Q-Q plots of these data and see if we have normal distributions. 

```{r t_test_qq, echo=FALSE}
qqnorm(non_rain$count, main='Q-Q Plot for Non-Precip')
qqnorm(rain_counts$count, main='Q-Q Plot for Precip')
```

In light of the Q-Q plots, it appears that both our Precip/Non-Precip samples come from normal populations to which we can perform a T-test on.  

### Welch T-Test
Now we can perform our Welch T-Test:
```{r t_test, echo=FALSE}
t.test(rain_counts$count,non_rain$count)
```

### Results
With a p-value < .05, we can reject **Hø**. The data are strong evidence that the population mean commuter counts are different on days of precipitation v. days of no precipitation. 

### Additional Questions
- Perform the same test for commuter countage during the academic year -- are college students more likely to ride in the rain? Do they have a choice? 
- Look at other variables such as temperature and wind speed.   
<br/>
<br/>
<br/>